---
title: "Assay vs Genes - Clean Verification"
author: "Diane"
date: "2026-03-03"
output: html_document
---

```{r setup, include=FALSE}
library(readxl)
library(dplyr)
```

# Import data
```{r import}
assay_genome_path <- "./data/Assay_vs_Genomes.xlsx"

df_assay_genome <- read_excel(assay_genome_path)

# Remove any fully empty rows
df_assay_genome <- df_assay_genome %>% filter(!is.na(isolate_number))

cat("Total isolates:", nrow(df_assay_genome), "\n")
```

# Define pathway gene sets and filter columns
```{r define_pathways}
meta_cols <- c(
  "Phylum", "Class", "Order", "Family", "Genus",
  "isolate_number", "GTDB_hit_iso_number",
  "N_fixing", "S_Oxid", "S_Red", "IAA_production", "P_solubility"
)

pathway_genes <- list(
  Nitrogen_fixation = c("K02586", "K02588", "K02591"),
  Denitrification = c("K00368", "K00370", "K00371", "K00374", "K00376",
                      "K02305", "K02567", "K02568", "K04561", "K15864"),
  DNRA = c("K00362", "K00363", "K03385", "K15876"),
  C_N_Bond_cleavage = c("K00274", "K01426", "K01501"),
  Sulfur_oxidation_SOX = c("K17222", "K17223", "K17224", "K17225", "K17226", "K17227", "K22622"),
  Sulfide_oxidation = c("K17229", "K17230"),
  Assimilatory_sulfur_reduction = c("K00380", "K00381", "K00392"),
  Thiosulfate_oxidation_non_SOX = c("K16936", "K16937", "K19713"),
  Phosphate_solubilizing = c("K01113", "K01077", "K00117"),
  Indole_Acetic_Acid = c("K00466", "K00832", "K04103", "K21801")
)

all_kegg_ids <- unlist(pathway_genes)
cols_to_keep <- intersect(c(meta_cols, all_kegg_ids), names(df_assay_genome))
df <- df_assay_genome[, cols_to_keep]

cat("Columns kept:", length(cols_to_keep), "\n")
```

# Helper function
```{r helper}
has_any_gene <- function(df, genes) {
  present <- intersect(genes, names(df))
  if (length(present) == 0) return(rep(FALSE, nrow(df)))
  rowSums(df[, present, drop = FALSE] == 1, na.rm = TRUE) > 0
}

has_all_genes <- function(df, genes) {
  present <- intersect(genes, names(df))
  if (length(present) == 0) return(rep(FALSE, nrow(df)))
  rowSums(df[, present, drop = FALSE] == 1, na.rm = TRUE) == length(present)
}
```

---------------
# NITROGEN
---------------

## Overall nitrogen gene counts
```{r nitrogen_overview}
Nitrogen_fixation <- c("K02586", "K02588", "K02591")
Denitrification <- c("K00368", "K00370", "K00371", "K00374", "K00376",
                    "K02305", "K02567", "K02568", "K04561", "K15864")
DNRA <- c("K00362", "K00363", "K03385", "K15876")
C_N_Bond_cleavage <- c("K00274", "K01426", "K01501")

df$has_N_fixation_genes <- has_any_gene(df, Nitrogen_fixation)
df$has_denitrification_genes <- has_any_gene(df, Denitrification)
df$has_DNRA_genes <- has_any_gene(df, DNRA)
df$has_C_N_bond_cleavage_genes <- has_any_gene(df, C_N_Bond_cleavage)

df$has_any_N_pathway_gene <- (
  df$has_N_fixation_genes |
  df$has_denitrification_genes |
  df$has_DNRA_genes |
  df$has_C_N_bond_cleavage_genes
)

total <- nrow(df)
cat("=== NITROGEN GENE COUNTS (all isolates) ===\n")
cat("N fixation (nifHDK):", sum(df$has_N_fixation_genes),
    sprintf("(%.1f%%)", sum(df$has_N_fixation_genes)/total*100),
    "| Manuscript: 10%\n")
cat("Denitrification:", sum(df$has_denitrification_genes),
    sprintf("(%.1f%%)", sum(df$has_denitrification_genes)/total*100),
    "| Manuscript: 51%\n")
cat("DNRA:", sum(df$has_DNRA_genes),
    sprintf("(%.1f%%)", sum(df$has_DNRA_genes)/total*100),
    "| Manuscript: 71%\n")
cat("C-N bond cleavage:", sum(df$has_C_N_bond_cleavage_genes),
    sprintf("(%.1f%%)", sum(df$has_C_N_bond_cleavage_genes)/total*100),
    "| Manuscript: 83%\n")
cat("ANY N mobilization:", sum(df$has_any_N_pathway_gene),
    sprintf("(%.1f%%)", sum(df$has_any_N_pathway_gene)/total*100),
    "| Manuscript: ~96%\n")
```

## N-fixing assay results vs genes
```{r nitrogen_assay}
cat("N_fixing assay positive:", sum(df$N_fixing == 1, na.rm = TRUE),
    "| Manuscript: 37\n\n")

n_fix_pos <- df %>% filter(N_fixing == 1)

summary_n <- n_fix_pos %>%
  summarize(
    total_N_fix_pos = n(),
    with_nifHDK = sum(has_N_fixation_genes),
    with_CN_bond = sum(has_C_N_bond_cleavage_genes),
    with_denitrification = sum(has_denitrification_genes),
    with_DNRA = sum(has_DNRA_genes),
    with_nifHDK_or_CN = sum(has_N_fixation_genes | has_C_N_bond_cleavage_genes),
    lacking_both_nif_and_CN = sum(!has_N_fixation_genes & !has_C_N_bond_cleavage_genes)
  )
print(summary_n)
```

## Identify isolates with nifHDK (robust growers per manuscript)
```{r nifHDK_isolates}
cat("=== Isolates with nifHDK genes ===\n")
cat("Manuscript says: 6 robust growers included Agarivorans sp. with nifHDK\n\n")

nifHDK_isolates <- df %>%
  filter(has_N_fixation_genes) %>%
  select(isolate_number, GTDB_hit_iso_number, Family, N_fixing)
print(nifHDK_isolates)
cat("\nCount:", nrow(nifHDK_isolates), "\n")
```

## NirBD verification for isolates lacking nifHDK and C:N bond cleavage
```{r nirBD_check}
# Manuscript says: Two isolates, Thalassospira lohafexi and Vibrio Splendidus,
# lacked nitrilase/amidase/MOA but contained assimilatory nitrite reductase (NirBD)
# NirBD = K00362 (nirB) + K00363 (nirD)

NirBD <- c("K00362", "K00363")
df$has_nirBD <- has_any_gene(df, NirBD)

cat("=== N_fixing+ isolates lacking BOTH nifHDK AND C:N bond cleavage genes ===\n")
cat("Manuscript says: 2 (Thalassospira lohafexi, Vibrio Splendidus) with NirBD\n\n")

n_fix_no_nif_no_cn <- df %>%
  filter(N_fixing == 1, !has_N_fixation_genes, !has_C_N_bond_cleavage_genes) %>%
  select(isolate_number, GTDB_hit_iso_number, Family,
         has_nirBD, has_DNRA_genes, has_denitrification_genes)

print(n_fix_no_nif_no_cn)

# Also show individual nirB/nirD presence
nirBD_detail <- df %>%
  filter(N_fixing == 1, !has_N_fixation_genes, !has_C_N_bond_cleavage_genes) %>%
  select(isolate_number, GTDB_hit_iso_number,
         K00362, K00363)  # nirB, nirD
print(nirBD_detail)
```

## N_fixing+ with C:N bond cleavage (nitrogen scavengers)
```{r cn_bond_scavengers}
# Manuscript says: 29 of the 31 limited-growth isolates coded for
# nitrilase (K01501), amidase amiE (K01426), and/or monoamine oxidase MOA (K00274)

n_fix_with_cn <- df %>%
  filter(N_fixing == 1, !has_N_fixation_genes, has_C_N_bond_cleavage_genes)

cat("N_fix+ with C:N bond cleavage (no nifHDK):", nrow(n_fix_with_cn),
    "| Manuscript: 29\n\n")

# Breakdown by individual gene
n_fix_no_nif <- df %>% filter(N_fixing == 1, !has_N_fixation_genes)
cat("Among N_fix+ without nifHDK (n=", nrow(n_fix_no_nif), "):\n")
cat("  with K01501 (nitrilase):", sum(n_fix_no_nif$K01501 == 1, na.rm = TRUE), "\n")
cat("  with K01426 (amidase amiE):", sum(n_fix_no_nif$K01426 == 1, na.rm = TRUE), "\n")
cat("  with K00274 (monoamine oxidase):", sum(n_fix_no_nif$K00274 == 1, na.rm = TRUE), "\n")
```

## Summary table for N_fixing positive isolates
```{r nitrogen_summary_table}
summary_simple <- df %>%
  filter(N_fixing == 1) %>%
  select(Class, Family, GTDB_hit_iso_number,
         has_N_fixation_genes, has_C_N_bond_cleavage_genes,
         has_denitrification_genes, has_DNRA_genes, has_nirBD) %>%
  mutate(
    Notes = case_when(
      has_N_fixation_genes ~ "Has nifHDK (robust grower)",
      has_C_N_bond_cleavage_genes ~ "N scavenger (nitrilase/amidase/MOA)",
      has_nirBD ~ "Has NirBD only",
      TRUE ~ "No known N genes"
    )
  )
print(summary_simple)
```

---------------
# SULFUR
---------------

## Overall sulfur gene counts
```{r sulfur_overview}
Sulfur_oxidation_SOX <- c("K17222", "K17223", "K17224", "K17225", "K17226", "K17227", "K22622")
Sulfide_oxidation <- c("K17229", "K17230")
Thiosulfate_oxidation_non_SOX <- c("K16936", "K16937", "K19713")
Assimilatory_sulfate_reduction <- c("K00380", "K00381", "K00392")

df$has_Sox_genes <- has_any_gene(df, Sulfur_oxidation_SOX)
df$has_sulfide_oxidation_genes <- has_any_gene(df, Sulfide_oxidation)
df$has_thiosulfate_oxidation_genes <- has_any_gene(df, Thiosulfate_oxidation_non_SOX)
df$has_assimilatory_sulfate_reduction <- has_any_gene(df, Assimilatory_sulfate_reduction)

# Sulfur/thiosulfate oxidation genes (SOX + sulfide + thiosulfate non-SOX)
all_sulfur_ox <- c(Sulfur_oxidation_SOX, Sulfide_oxidation, Thiosulfate_oxidation_non_SOX)
df$has_any_sulfur_ox_gene <- has_any_gene(df, all_sulfur_ox)

# Any sulfur gene
df$has_any_sulfur_gene <- (
  df$has_Sox_genes | df$has_sulfide_oxidation_genes |
  df$has_thiosulfate_oxidation_genes | df$has_assimilatory_sulfate_reduction
)

cat("=== SULFUR GENE COUNTS (all isolates) ===\n")
cat("With SOX genes:", sum(df$has_Sox_genes), "\n")
cat("With sulfide oxidation:", sum(df$has_sulfide_oxidation_genes), "\n")
cat("With thiosulfate non-SOX:", sum(df$has_thiosulfate_oxidation_genes), "\n")
cat("With assimilatory sulfate reduction (cysI/cysJ/sir):", sum(df$has_assimilatory_sulfate_reduction), "\n")
cat("With any sulfur/thiosulfate OXIDATION gene:", sum(df$has_any_sulfur_ox_gene),
    sprintf("(%.1f%%)", sum(df$has_any_sulfur_ox_gene)/total*100),
    "| Manuscript: 52%\n")
cat("With any sulfur gene:", sum(df$has_any_sulfur_gene), "\n")
```

## Sulfur assay results
```{r sulfur_assay}
cat("S_Oxid positive (acidified):", sum(df$S_Oxid == 1, na.rm = TRUE),
    "| Manuscript: 9\n")
cat("S_Red positive:", sum(df$S_Red == 1, na.rm = TRUE), "\n")
cat("S_Oxid OR S_Red:", sum(df$S_Oxid == 1 | df$S_Red == 1, na.rm = TRUE),
    "| Manuscript: 41 grew on thiosulfate\n")
```

## Complete soxABCXYZ check
```{r complete_sox}
# Manuscript says: only Roseibium sp. (iso195) had the complete set (soxABCXYZ)
# Check which SOX genes are actually in the data
sox_genes_available <- intersect(Sulfur_oxidation_SOX, names(df))
cat("SOX genes available in data:", paste(sox_genes_available, collapse=", "), "\n")
cat("SOX genes NOT in data:", paste(setdiff(Sulfur_oxidation_SOX, names(df)), collapse=", "), "\n\n")

df$has_complete_SOX <- has_all_genes(df, Sulfur_oxidation_SOX)

cat("=== Isolates with COMPLETE SOX complex ===\n")
cat("Manuscript says: only Roseibium sp. (iso195)\n\n")

complete_sox <- df %>%
  filter(has_complete_SOX) %>%
  select(isolate_number, GTDB_hit_iso_number, Family)
print(complete_sox)
cat("Count:", nrow(complete_sox), "\n")
```

## Verify cysI/cysJ/sir among thiosulfate growers
```{r cysI_cysJ_sir_check}
# Manuscript says: All 41 isolates, except Rossellomorea sp. (iso230),
# had sulfite reductase [NADPH] (cysI, cysJ) or sulfite reductase ferredoxin (sir)

thio_growers <- df %>% filter(S_Oxid == 1 | S_Red == 1)

cat("=== Thiosulfate growers WITHOUT cysI/cysJ/sir ===\n")
cat("Manuscript says: only Rossellomorea sp. (iso230)\n\n")

thio_no_assim <- thio_growers %>%
  filter(!has_assimilatory_sulfate_reduction) %>%
  select(isolate_number, GTDB_hit_iso_number, Family, S_Oxid, S_Red)
print(thio_no_assim)
cat("Count without cysI/cysJ/sir:", nrow(thio_no_assim), "\n")
cat("Count with cysI/cysJ/sir:", nrow(thio_growers) - nrow(thio_no_assim), "\n")
```

## S_Oxid positive (acidified) isolates detail
```{r sulfur_oxid_detail}
cat("=== S_Oxid positive isolates and families ===\n")
cat("Manuscript families: Micrococcaceae, Streptomycetaceae, Bacillaceae,\n")
cat("  Alteromonadaceae, Stappiaceae, Vibrionaceae\n\n")

s_oxid_pos <- df %>%
  filter(S_Oxid == 1) %>%
  select(isolate_number, GTDB_hit_iso_number, Family,
         has_Sox_genes, has_sulfide_oxidation_genes,
         has_thiosulfate_oxidation_genes, has_assimilatory_sulfate_reduction,
         has_complete_SOX)
print(s_oxid_pos)

cat("\nFamilies of S_Oxid positive isolates:\n")
s_oxid_families <- df %>% filter(S_Oxid == 1) %>% count(Family, sort = TRUE)
print(s_oxid_families)
```

## Sulfur phenotype summary table
```{r sulfur_summary_table}
sulfur_pheno_pos <- df %>%
  filter(S_Oxid == 1 | S_Red == 1) %>%
  select(Class, Family, GTDB_hit_iso_number,
         S_Oxid, S_Red,
         has_Sox_genes, has_sulfide_oxidation_genes,
         has_thiosulfate_oxidation_genes, has_assimilatory_sulfate_reduction,
         has_complete_SOX) %>%
  mutate(
    Notes = case_when(
      has_complete_SOX ~ "Complete SOX complex",
      !(has_Sox_genes | has_sulfide_oxidation_genes |
        has_thiosulfate_oxidation_genes | has_assimilatory_sulfate_reduction) ~
        "Phenotype positive, lacks sulfur genes",
      TRUE ~ ""
    )
  )
print(sulfur_pheno_pos)
```

## Phenotype positive but lacking sulfur genes
```{r sulfur_phenotype_no_genes}
phenotype_pos_no_genes <- df %>%
  filter(
    (S_Oxid == 1 | S_Red == 1) &
    !(has_Sox_genes | has_sulfide_oxidation_genes |
      has_thiosulfate_oxidation_genes | has_assimilatory_sulfate_reduction)
  ) %>%
  select(isolate_number, GTDB_hit_iso_number, S_Oxid, S_Red)
cat("Sulfur phenotype positive but lacking all screened genes:\n")
print(phenotype_pos_no_genes)
```

## Phenotype negative but with sulfur genes
```{r sulfur_neg_with_genes}
phenotype_neg_with_genes <- df %>%
  filter(
    S_Oxid == 0 & S_Red == 0 &
    (has_Sox_genes | has_sulfide_oxidation_genes |
     has_thiosulfate_oxidation_genes | has_assimilatory_sulfate_reduction)
  ) %>%
  select(isolate_number, GTDB_hit_iso_number,
         has_Sox_genes, has_sulfide_oxidation_genes,
         has_thiosulfate_oxidation_genes, has_assimilatory_sulfate_reduction)
cat("Sulfur phenotype negative but with sulfur genes:\n")
print(phenotype_neg_with_genes)
```

---------------
# PHOSPHORUS
---------------

## Overall phosphorus gene counts
```{r phosphorus_overview}
p_genes <- c("K01113", "K01077", "K00117")  # phoD, phoA, gcd
df$has_any_P_gene <- has_any_gene(df, p_genes)

cat("=== PHOSPHORUS GENE COUNTS (all isolates) ===\n")
cat("With any P-solubilizing gene:", sum(df$has_any_P_gene),
    sprintf("(%.1f%%)", sum(df$has_any_P_gene)/total*100),
    "| Manuscript: 88.5%\n")
cat("  phoD (K01113):", sum(df$K01113 == 1, na.rm = TRUE), "\n")
cat("  phoA (K01077):", sum(df$K01077 == 1, na.rm = TRUE), "\n")
cat("  gcd  (K00117):", sum(df$K00117 == 1, na.rm = TRUE), "\n")
```

## P_solubility assay vs genes
```{r phosphorus_assay}
cat("P_solubility positive:", sum(df$P_solubility == 1, na.rm = TRUE), "\n\n")

p_sol_pos <- df %>% filter(P_solubility == 1)

cat("Among P_solubility positive isolates:\n")
cat("  with phoD (K01113):", sum(p_sol_pos$K01113 == 1, na.rm = TRUE), "\n")
cat("  with phoA (K01077):", sum(p_sol_pos$K01077 == 1, na.rm = TRUE), "\n")
cat("  with gcd  (K00117):", sum(p_sol_pos$K00117 == 1, na.rm = TRUE), "\n")
cat("  with any P gene:", sum(has_any_gene(p_sol_pos, p_genes)), "\n")
cat("  with both phoD and phoA:", sum(p_sol_pos$K01113 == 1 & p_sol_pos$K01077 == 1, na.rm = TRUE), "\n")
```

## P phenotype positive but lacking genes
```{r phosphorus_no_genes}
p_pos_no_genes <- df %>%
  filter(P_solubility == 1 & !has_any_P_gene) %>%
  select(isolate_number, GTDB_hit_iso_number, P_solubility, K01113, K01077, K00117)
cat("P_solubility positive but lacking all P genes:\n")
print(p_pos_no_genes)
```

## P phenotype negative but with genes
```{r phosphorus_neg_with_genes}
p_neg_with_genes <- df %>%
  filter(P_solubility == 0 & has_any_P_gene) %>%
  select(isolate_number, GTDB_hit_iso_number, K01113, K01077, K00117)
cat("P_solubility negative but with P genes:\n")
print(p_neg_with_genes)

cat("\nFamilies of P_sol negative with P genes:\n")
df %>%
  filter(P_solubility == 0 & has_any_P_gene) %>%
  count(Family, sort = TRUE) %>%
  print()
```

## Family-level P_solubility discordance
```{r phosphorus_family}
family_summary <- df %>%
  group_by(Family) %>%
  summarise(
    n_total = n(),
    n_pos = sum(P_solubility == 1, na.rm = TRUE),
    n_neg_with_genes = sum(P_solubility == 0 & has_any_P_gene, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_pos > 0 & n_neg_with_genes > 0)
cat("Families with both P+ isolates and P- isolates carrying P genes:\n")
print(family_summary)
```

---------------
# IAA (Indole-3-Acetic Acid)
---------------

## Overall IAA gene counts
```{r iaa_overview}
iaa_genes <- c("K00466", "K21801", "K04103", "K00832")  # iaaM, iaaH, ipdC, tryB
df$has_any_iaa_gene <- has_any_gene(df, iaa_genes)

cat("=== IAA GENE COUNTS (all isolates) ===\n")
cat("With any IAA gene:", sum(df$has_any_iaa_gene),
    sprintf("(%.1f%%)", sum(df$has_any_iaa_gene)/total*100),
    "| Manuscript: 60.5%\n")
cat("  iaaM  (K00466):", sum(df$K00466 == 1, na.rm = TRUE), "\n")
cat("  iaaH  (K21801):", sum(df$K21801 == 1, na.rm = TRUE), "\n")
cat("  ipdC  (K04103):", sum(df$K04103 == 1, na.rm = TRUE), "\n")
cat("  tryB  (K00832):", sum(df$K00832 == 1, na.rm = TRUE), "\n")
```

## IAA assay vs genes
```{r iaa_assay}
cat("IAA_production positive:", sum(df$IAA_production == 1, na.rm = TRUE), "\n\n")

iaa_pos <- df %>% filter(IAA_production == 1)

cat("Among IAA+ isolates:\n")
cat("  with any IAA gene:", sum(has_any_gene(iaa_pos, iaa_genes)), "\n")
cat("  without any IAA gene:", sum(!has_any_gene(iaa_pos, iaa_genes)), "\n")
cat("  with K00832 (tryB):", sum(iaa_pos$K00832 == 1, na.rm = TRUE), "\n")

iaa_pos_detail <- iaa_pos %>%
  select(GTDB_hit_iso_number, IAA_production, K00466, K21801, K04103, K00832)
print(iaa_pos_detail)
```

## IAA positive but lacking genes
```{r iaa_no_genes}
iaa_pos_no_genes <- df %>%
  filter(IAA_production == 1 & !has_any_iaa_gene) %>%
  select(isolate_number, GTDB_hit_iso_number, K00466, K21801, K04103, K00832)
cat("IAA+ but lacking all screened IAA genes:\n")
print(iaa_pos_no_genes)
```

## IAA negative but with genes
```{r iaa_neg_with_genes}
iaa_neg_with_genes <- df %>%
  filter(IAA_production == 0 & has_any_iaa_gene) %>%
  select(isolate_number, GTDB_hit_iso_number, K00466, K21801, K04103, K00832)
cat("IAA- but with IAA genes:\n")
print(iaa_neg_with_genes)
```

---------------
# Other PGP genes
---------------

```{r other_pgp}
cat("=== OTHER PGP GENES (all isolates) ===\n")
cat("arsR (K03892) arsenic degradation:", sum(df_assay_genome$K03892 == 1, na.rm = TRUE), "\n")
cat("ribBA (K14652) folate biosynthesis:", sum(df_assay_genome$K14652 == 1, na.rm = TRUE), "\n")
cat("corA (K03284) cobalt transporter:", sum(df_assay_genome$K03284 == 1, na.rm = TRUE), "\n")
```

## Gibberellin and terpenoid precursors
```{r gibberellin}
cat("idsA (K13787) geranylgeranyl diphosphate synthase type I:",
    sum(df_assay_genome$K13787 == 1, na.rm = TRUE), "\n")
cat("ferredoxin (K05337) gibberellin biosynthesis:",
    sum(df_assay_genome$K05337 == 1, na.rm = TRUE), "\n")
cat("GGPS (K13789) geranylgeranyl diphosphate synthase:",
    sum(df_assay_genome$K13789 == 1, na.rm = TRUE), "\n")
```

## Glyoxylate metabolism
```{r glyoxylate}
cat("=== GLYOXYLATE ===\n")
cat("ghrA (K12972) glyoxylate/hydroxypyruvate reductase:",
    sum(df_assay_genome$K12972 == 1, na.rm = TRUE), "\n")
cat("mch/mcd (K14449) 2-methylfumaryl-CoA hydratase:",
    sum(df_assay_genome$K14449 == 1, na.rm = TRUE), "\n")
cat("aceA (K01637) isocitrate lyase:",
    sum(df_assay_genome$K01637 == 1, na.rm = TRUE), "\n")
cat("glcB (K01638) malate synthase:",
    sum(df_assay_genome$K01638 == 1, na.rm = TRUE), "\n")
```

---------------
# Secretion Systems
---------------

```{r secretion_setup}
secretion_systems <- list(
  "Sec-SRP pathway" = c("K03072", "K03073", "K03074", "K03075", "K03076", "K03210", "K03217"),
  "Twin Arginine Targeting (Tat)" = c("K03116", "K03117", "K03118", "K03425"),
  "Type I Secretion System" = c("K12340", "K11003", "K11004"),
  "Type II Secretion System" = c("K02453", "K02465", "K02455", "K02457", "K02458", "K02459", "K02460", "K02461", "K02462"),
  "Type III Secretion System" = c("K03221", "K04056", "K04057", "K04059", "K03227", "K03228", "K03229", "K03230"),
  "Type IV Secretion System" = c("K03194", "K03197", "K03198", "K03200", "K03202", "K03204"),
  "Type V Secretion System (Autotransporter)" = c("K11028", "K12341", "K12342"),
  "Type VI Secretion System" = c("K11891", "K11892", "K11904", "K11903", "K11906", "K11907", "K11912")
)

# Need to re-filter df to include secretion KOs
secretion_kos <- unlist(secretion_systems, use.names = FALSE)
sec_cols_to_keep <- intersect(c(meta_cols, all_kegg_ids, secretion_kos), names(df_assay_genome))
df_sec <- df_assay_genome[!is.na(df_assay_genome$isolate_number), sec_cols_to_keep]

for (sys_name in names(secretion_systems)) {
  kos <- secretion_systems[[sys_name]]
  present_kos <- intersect(kos, names(df_sec))
  col_name <- paste0("has_any_", gsub("[() ]", "_", tolower(sys_name)))
  df_sec[[col_name]] <- has_any_gene(df_sec, present_kos)
}
```

```{r secretion_counts}
cat("=== SECRETION SYSTEM COUNTS ===\n")
for (sys_name in names(secretion_systems)) {
  col_name <- paste0("has_any_", gsub("[() ]", "_", tolower(sys_name)))
  if (col_name %in% names(df_sec)) {
    cat(sys_name, ":", sum(df_sec[[col_name]], na.rm = TRUE), "\n")
  }
}
```

```{r secretion_detail}
for (sys_name in names(secretion_systems)) {
  col_name <- paste0("has_any_", gsub("[() ]", "_", tolower(sys_name)))
  if (col_name %in% names(df_sec)) {
    pos <- df_sec %>% filter(.data[[col_name]])
    if (nrow(pos) > 0) {
      gene_cols <- intersect(secretion_systems[[sys_name]], names(df_sec))
      cat("\n===", toupper(sys_name), "- positive isolates:", nrow(pos), "===\n")
      output <- pos %>% select(isolate_number, GTDB_hit_iso_number, all_of(gene_cols))
      print(output)
    }
  }
}
```
